Fast time-domain electromagnetic modelling in the frequency domain

### 2020-11-24 ManuscriptTime.tex
Change from Geophysics to GJI template

\documentclass[
%     paper,
    manuscript,
%     twocolumn,
%     twoside,
    revised,
  ]{geophysics}

% Remove once done
\usepackage[textsize=footnotesize, textwidth=2.8cm]{todonotes}

% Additional packages to geophysics.cls
\usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}

\usepackage{amssymb, amsmath, amsfonts}
\usepackage{upquote}
\usepackage[strings]{underscore}
\usepackage{siunitx}                % SI conform commands (eg \num)
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{xspace}

\usepackage[pdftex, final]{hyperref}
% \usepackage[pdftex, hidelinks]{hyperref}
\hypersetup{allcolors=blue, allbordercolors={0 0 .5}, colorlinks=true}

% Figures
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\renewcommand{\figdir}{./figures}
\ifthenelse{\boolean{@twoc}}{%
  \newcommand{\cwidth}{250pt}}{%
  \newcommand{\cwidth}{240pt}}

\ifthenelse{\boolean{@twoc}}{%
  \newcommand{\acwidth}{250pt}%
  \newcommand{\tcwidth}{250pt}}{%
  \newcommand{\acwidth}{240pt}%
  \newcommand{\tcwidth}{312pt}}

@@

@@

\title{Fast time-domain electromagnetic modelling in the frequency domain}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\ms{}  % ARTICLE-ID

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]Shell Global Solutions International BV,
Grasweg 31, 1031 HW Amsterdam;\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
}

\author{%
Dieter Werthmüller\footnotemark[1], %            orcid: 0000-0002-8575-2484
Wim A.\ Mulder\footnotemark[1]\footnotemark[2], % orcid: 0000-0001-7020-9297
and Evert C.\ Slob\footnotemark[1] %             orcid: 0000-0002-4529-1134
}

\footer{}
\lefthead{Werthmüller et al.}
\righthead{Time-domain CSEM}

\maketitle

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. The fast
frequency-to-time domain method uses either the digital linear filter method
or the logarithmic fast Fourier transform together with a careful selection of
frequencies and interpolation. This methodology requires typically 15 to 20
frequencies to cover a wide range of offsets in one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three \new{key} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and the corresponding grids
are carefully chosen.
%
\\[1cm] \textbf{Keywords}: Controlled source electromagnetics (CSEM); Numerical
modelling; Fourier analysis.
\end{abstract}

\section{Introduction}

@@

\section{Motivation}

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains\old{, as it is done, for instance, in \simpeg (Cockett et al.,
2015)}. Another approach is to have Maxwell's equation only implemented in one
domain, and use Fourier transforms to go to the other. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, translating into a
significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

\section{Methodology}

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \old{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \new{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu 18.04 and Python 3.7. For the
semi-analytical computations of layered models we use the open-source (Apache
License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\section{Frequency selection}

@@

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\plot{GUI-FFTLog-lin}{width=\tcwidth}{
  Interactive frequency selection for a user-provided layered model (the shown
  model parameters are described in the text). This example shows the impulse
  response at an offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to
  10\,Hz with five frequencies per decade. Note that the error is clipped for
  values smaller than 0.01\,\% and bigger than 100\,\%.
}
%
The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m at 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the \new{semi-analytical
responses}\old{precise results obtained from \empymod}. The blue circles
indicate the actually computed responses and the black dots the frequencies
which are interpolated or set to zero. The resulting time-domain response has a
relative error of less than 1\,\% everywhere except for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate. Figure \ref{fig:GUI-DLF-lin} is the same as
\ref{fig:GUI-FFTLog-lin}, but transformed with the DLF method applying the
81-point sine-cosine filter from \cite{GEO.09.Key}. The same frequencies were
computed as in the FFTLog case and the missing ones interpolated. The error of
the corresponding time-domain response is comparable, so either FFTLog or DLF
can be used.
%
\multiplot{2}{GUI-FFTLog-log,GUI-DLF-lin}{width=\tcwidth}
{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic scale. (b)
Using DLF instead of FFTLog, but computing the same frequenc\old{ies}\new{y
range} as for the FFTLog. The resulting time domain response has comparable
accuracy (compare the error to the error in Figure \ref{fig:GUI-FFTLog-lin}).
}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. However, with this frequency-selection we can
model a wide range of offsets. This is shown in Figure \ref{fig:multi-offset},
where the above parameters were used to model the responses at offsets
$r=1.5,3,6$, and 12\,km.
%
\plot{multi-offset}{width=\tcwidth}{
  Normalized (a) frequency- and (b) time-domain responses for offsets
  $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
  \ref{fig:GUI-FFTLog-lin}. The coloured circles are the actually computed
  responses, the black dots are the responses which are set to zero or
  interpolated. The black curves are the analytical responses.
}
%

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the \old{less}\new{fewer} frequencies are needed.
Another important factor is how to interpolate and extrapolate from the
computed frequencies to the frequencies required for the Fourier transform. For
the FFTLog only extrapolation for higher and lower frequencies is required. The
EM response becomes highly oscillatory for high frequencies, which makes it
very hard to extrapolate the response to frequencies $f>f_\mathrm{max}$.
However, if $f_\mathrm{max}$ is chosen judiciously, the importance of higher
frequencies for the Fourier transform can be neglected and we can set those
responses to zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be
changed to an interpolation by assuming a zero imaginary response at zero
frequency, and we then use PCHIP to interpolate the missing frequencies (as we
work on a logarithmic scale we cannot choose 0\,Hz, but instead take
$10^{-100}$\,Hz). In the case of the DLF method we also have to interpolate in
between the computed frequencies, for which we found it better to use a cubic
spline. As can be seen from Figures \ref{fig:GUI-FFTLog-lin} and
\ref{fig:GUI-DLF-lin}, using the FFTLog or the DLF with the same actually
computed frequencies results in very similar responses.

@@

\section{Gridding}

@@

In summary, the adaptive gridding takes $f$, $D_\mr{s}$, $\sigma_\mr{src}$,
$\sigma_\mr{ave}$, $n_\delta$, and ranges for $\alpha_\mr{s}$, $\alpha_\mr{c}$,
where we usually fix $\alpha_\mr{s}=1$ or keep it at least below 1.05, and let
$\alpha_\mr{c}$ be anything between $[1, 1.5]$. The minimum cell width
$\Delta_\mr{min}$ can further be restricted by a user-defined range. Given
these inputs the adaptive gridding will search for the smallest possible number
of cells which fulfils these criteria. The \new{implemented} multigrid method
\old{implemented in \emg3d} puts some constraints on the number of cells, of
which the adaptive gridding takes care (the number of cells have to be powers
of two multiplied by a low prime, e.g., $\{2,3,5\}\cdot2^n$).

@@

\section{Numerical examples}

\subsection{Homogeneous space}

The first example is the inline electric field from a source at the origin
measured by an inline receiver with an offset of 900\,m in a homogeneous space
of $1\,\Omega\,$m. We chose this simple example to compare it with the
analytical solution and with previously published results. We used the
following values to define the required frequencies: $f_\mr{min}=0.05\,$Hz,
$f_\mr{max}=21\,$Hz, using FFTLog with 5 frequencies per decade. This results
in 14 frequencies to compute from 0.05\,Hz to 20.0\,Hz. The complete frequency
range for the transform includes 30 frequencies from 0.0002\,Hz to 126.4\,Hz.
For the adaptive gridding the following inputs were used: $n_\delta=12$,
minimum cell width must be between 20 and 40\,m, and $\alpha_\mr{s}=1$,
$\alpha_\mr{c}=[1,1.3]$. This created grids with cell numbers between
\num{46080} ($80\times24\times24$, for 20.0\,Hz) and \num{128000}
($80\times40\times40$, for 0.05\,Hz) cells. The run times for each frequency,
the corresponding number of cells, minimum cell width and computation domain
stretching factor are listed in Table~\ref{tbl:timefull}. The total run time to
compute this model was less than two minutes.
%
\tabl[btp]{timefull}{Run times per frequency for the homogeneous space example,
  with the corresponding number of cells and minimum cell width as well as the
  stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}{
  \centering
\begin{tabular}{S[table-format=2.4]rrrr}
  \toprule
  %
  {freq.} & time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
  {(Hz)}  & (s) &  & (m) & \\
  \midrule
  %
  20.0    &  3 & 80$\times$24$\times$24 & 20 & 1.26 \\
  12.6    &  6 & 96$\times$32$\times$32 & 20 & 1.17 \\
   7.98   &  8 & 96$\times$32$\times$32 & 20 & 1.20 \\
   5.03   &  7 & 96$\times$32$\times$32 & 20 & 1.23 \\
   3.18   &  7 & 80$\times$32$\times$32 & 24 & 1.25 \\
   2.00   &  5 & 80$\times$32$\times$32 & 30 & 1.21 \\
   1.26   &  4 & 64$\times$32$\times$32 & 37 & 1.21 \\
   0.798  &  5 & 64$\times$32$\times$32 & 40 & 1.23 \\
   0.503  &  5 & 64$\times$32$\times$32 & 40 & 1.26 \\
   0.318  &  5 & 64$\times$32$\times$32 & 40 & 1.28 \\
   0.200  & 10 & 64$\times$40$\times$40 & 40 & 1.27 \\
   0.126  & 10 & 64$\times$40$\times$40 & 40 & 1.30 \\
   0.0798 & 14 & 80$\times$40$\times$40 & 40 & 1.26 \\
   0.0503 & 15 & 80$\times$40$\times$40 & 40 & 1.28 \\
  %
  \bottomrule
\end{tabular}}%
%

>>>>

% \documentclass[onecolumn,extra,referee,camera]{gji}
% \documentclass[onecolumn,extra,camera]{gji}
\documentclass[twocolumn,extra,camera]{gji}
\usepackage{timet}

% Additional packages
% \usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{amssymb, amsmath, amsfonts}
\usepackage{booktabs}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage[draft, addedmarkup=bf]{changes}
% \usepackage[final]{changes}
\usepackage[pdftex, final, allcolors=blue, colorlinks=true]{hyperref}

% workaround siunitx clashes unfortunately with gji
% https://tex.stackexchange.com/a/412886
\makeatletter
\let\zz@tabular\@tabular
\let\zzendtabular\endtabular
\let\zz@xtabularcr\@xtabularcr
\let\zz@tabclassz\@tabclassz
\let\zz@tabclassiv \@tabclassiv 
\let\zz@tabarray\@tabarray
\makeatother

% Change default colour of «changes» package
\definecolor{myred}{rgb}{.8,0,0}
\definecolor{mygreen}{rgb}{0,.5,0}
\setaddedmarkup{\color{myred}{#1}}
\setdeletedmarkup{\color{mygreen}{\itshape \sout{#1}}}

\let\tablenum\relax
\usepackage{siunitx}
\usepackage{tabularx}

% To fix clash of gji with hyperref
% https://tex.stackexchange.com/a/384969
\AtBeginDocument{
  \label{CorrectFirstPageLabel}
  \def\fpage{\pageref{CorrectFirstPageLabel}}
}

@@

\title[Time-domain CSEM]{Fast time-domain electromagnetic modelling in the frequency domain}

\author[D. Werthmüller \emph{et al.}]
  {\Large Dieter Werthmüller$^1$,        % orcid: 0000-0002-8575-2484
   Wim A. Mulder$^{1,2}$ and          % orcid: 0000-0001-7020-9297
   Evert C. Slob$^1$              % orcid: 0000-0002-4529-1134
   \\
   \footnotesize
  $^1$ TU Delft, Building 23, Stevinweg 1 / PO-box 5048, 2628 CN Delft (NL).
  E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org}\\[-.3em]
   \footnotesize
  $^2$ Shell Global Solutions International BV, Grasweg 31, 1031 HW Amsterdam(NL)
  }

\date{Received 202Y MMMM DD; in original form 2020 June 30}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{VVV}
\pubyear{202Y}

% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

\newtheorem{theorem}{Theorem}[section]

@@

\label{firstpage}

% workaround siunitx clashes unfortunately with gji
% https://tex.stackexchange.com/a/412886
{\makeatletter
\let\@tabular\zz@tabular
\let\endtabular\zzendtabular
\let\@xtabularcr\zz@xtabularcr
\let\@tabclassz\zz@tabclassz
\let\@tabclassiv \zz@tabclassiv 
\let\@tabarray\zz@tabarray
\maketitle
}
% \maketitle

\begin{summary}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. The fast
frequency-to-time domain method uses either the digital linear filter method
or the logarithmic fast Fourier transform together with a careful selection of
frequencies and interpolation. This methodology requires typically 15 to 20
frequencies to cover a wide range of offsets in one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three \added{key} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and the corresponding grids
are carefully chosen.
%
\end{summary}

% https://academic.oup.com/DocumentLibrary/GJI/gji%20keyword%20list%20updated2017.pdf
\begin{keywords}
  Controlled source electromagnetics (CSEM); Numerical modelling; Fourier analysis.
\end{keywords}

\section{Introduction}  % % % % % % % % % % % % % % % % % % % % % % % % % % % %

@@

\section{Motivation}  % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains\deleted{, as it is done, for instance, in \simpeg (Cockett et al.,
2015)}. Another approach is to have Maxwell's equation only implemented in one
domain, and use Fourier transforms to go to the other. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, translating into a
significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

\section{Methodology} % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu 18.04 and Python 3.7. For the
semi-analytical computations of layered models we use the open-source (Apache
License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\section{Frequency selection} % % % % % % % % % % % % % % % % % % % % % % % % %

@@

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/GUI-FFTLog-lin}
  \caption{Interactive frequency selection for a user-provided layered model
    (the shown model parameters are described in the text). This example shows
    the impulse response at an offset of 5\,km, for which it uses FFTLog from
    0.001\,Hz to 10\,Hz with five frequencies per decade. Note that the error
    is clipped for values smaller than 0.01\,\% and bigger than 100\,\%.}
  \label{fig:GUI-FFTLog-lin}
\end{figure}
%

The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m at 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the
\replaced{semi-analytical responses}{precise results obtained from \empymod}.
The blue circles indicate the actually computed responses and the black dots
the frequencies which are interpolated or set to zero. The resulting
time-domain response has a relative error of less than 1\,\% everywhere except
for very early times.

Figure \ref{fig:GUI-FFTLog-log} shows exactly the same on a logarithmic scale,
without the interactive GUI-related elements. It can be seen that with the
chosen frequencies the time-domain response starts to divert above about 100\,s
and below 0.3\,s. It also shows the oscillating high-frequency part, which is
hard to interpolate. Figure \ref{fig:GUI-DLF-lin} is the same as
\ref{fig:GUI-FFTLog-lin}, but transformed with the DLF method applying the
81-point sine-cosine filter from \cite{GEO.09.Key}. The same frequencies were
computed as in the FFTLog case and the missing ones interpolated. The error of
the corresponding time-domain response is comparable, so either FFTLog or DLF
can be used.
%
\begin{figure*}
  \centering
  \subfloat[]{{\includegraphics[width=.5\linewidth]{figures/GUI-FFTLog-log}}%
  \label{fig:GUI-FFTLog-log}}\hfill
  \subfloat[]{{\includegraphics[width=.5\linewidth]{figures/GUI-DLF-lin}}%
  \label{fig:GUI-DLF-lin}}\\
  \caption{(a) Same as Figure \ref{fig:GUI-FFTLog-lin}, but on a logarithmic
  scale. (b) Using DLF instead of FFTLog, but computing the same
  frequenc\replaced{y range}{ies} as for the FFTLog. The resulting time domain
  response has comparable accuracy (compare the error to the error in Figure
  \ref{fig:GUI-FFTLog-lin}).}
  \label{fig:GUI-log-lin}
\end{figure*}
%
In the above example we used 20 frequencies, but that many would not be
required for the shown offset of 5\,km, a few of the lower and higher
frequencies could be left out. However, with this frequency-selection we can
model a wide range of offsets. This is shown in Figure \ref{fig:multi-offset},
where the above parameters were used to model the responses at offsets
$r=1.5,3,6$, and 12\,km.
%
\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{figures/multi-offset}
  \caption{Normalized (a) frequency- and (b) time-domain responses for offsets
    $r=1.5,3,6$, and 12\,km using the parameters defined in Figure
    \ref{fig:GUI-FFTLog-lin}. The coloured circles are the actually computed
    responses, the black dots are the responses which are set to zero or
    interpolated. The black curves are the analytical responses.}
  \label{fig:multi-offset}
\end{figure}
%

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the \replaced{fewer}{less} frequencies are needed.
Another important factor is how to interpolate and extrapolate from the
computed frequencies to the frequencies required for the Fourier transform. For
the FFTLog only extrapolation for higher and lower frequencies is required. The
EM response becomes highly oscillatory for high frequencies, which makes it
very hard to extrapolate the response to frequencies $f>f_\mathrm{max}$.
However, if $f_\mathrm{max}$ is chosen judiciously, the importance of higher
frequencies for the Fourier transform can be neglected and we can set those
responses to zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be
changed to an interpolation by assuming a zero imaginary response at zero
frequency, and we then use PCHIP to interpolate the missing frequencies (as we
work on a logarithmic scale we cannot choose 0\,Hz, but instead take
$10^{-100}$\,Hz). In the case of the DLF method we also have to interpolate in
between the computed frequencies, for which we found it better to use a cubic
spline. As can be seen from Figures \ref{fig:GUI-FFTLog-lin} and
\ref{fig:GUI-DLF-lin}, using the FFTLog or the DLF with the same actually
computed frequencies results in very similar responses.

@@

\section{Gridding}  % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

@@

In summary, the adaptive gridding takes $f$, $D_\mr{s}$, $\sigma_\mr{src}$,
$\sigma_\mr{ave}$, $n_\delta$, and ranges for $\alpha_\mr{s}$, $\alpha_\mr{c}$,
where we usually fix $\alpha_\mr{s}=1$ or keep it at least below 1.05, and let
$\alpha_\mr{c}$ be anything between $[1, 1.5]$. The minimum cell width
$\Delta_\mr{min}$ can further be restricted by a user-defined range. Given
these inputs the adaptive gridding will search for the smallest possible number
of cells which fulfils these criteria. The \added{implemented} multigrid method
\deleted{implemented in \emg3d} puts some constraints on the number of cells, of
which the adaptive gridding takes care (the number of cells have to be powers
of two multiplied by a low prime, e.g., $\{2,3,5\}\cdot2^n$).

@@

\section{Numerical examples}  % % % % % % % % % % % % % % % % % % % % % % % % %

\subsection{Homogeneous space}  % % % % % % % % % % % % % % % % % % % % % % % %

The first example is the inline electric field from a source at the origin
measured by an inline receiver with an offset of 900\,m in a homogeneous space
of $1\,\Omega\,$m. We chose this simple example to compare it with the
analytical solution and with previously published results. We used the
following values to define the required frequencies: $f_\mr{min}=0.05\,$Hz,
$f_\mr{max}=21\,$Hz, using FFTLog with 5 frequencies per decade. This results
in 14 frequencies to compute from 0.05\,Hz to 20.0\,Hz. The complete frequency
range for the transform includes 30 frequencies from 0.0002\,Hz to 126.4\,Hz.
For the adaptive gridding the following inputs were used: $n_\delta=12$,
minimum cell width must be between 20 and 40\,m, and $\alpha_\mr{s}=1$,
$\alpha_\mr{c}=[1,1.3]$. This created grids with cell numbers between
\num{46080} ($80\times24\times24$, for 20.0\,Hz) and \num{128000}
($80\times40\times40$, for 0.05\,Hz) cells. The run times for each frequency,
the corresponding number of cells, minimum cell width and computation domain
stretching factor are listed in Table~\ref{tbl:timefull}. The total run time to
compute this model was less than two minutes.
%
\begin{table}
\begin{minipage}{\columnwidth}
  \centering
  \caption{Run times per frequency for the homogeneous space example, with the
    corresponding number of cells and minimum cell width as well as the
    stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}
  \label{tbl:timefull}
  \begin{tabular}{S[table-format=2.4]rrrr}
    \toprule
    %
    {freq.} & time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
    {(Hz)}  & (s) &  & (m) & \\
    \midrule
    %
    20.0    &  3 & 80$\times$24$\times$24 & 20 & 1.26 \\
    12.6    &  6 & 96$\times$32$\times$32 & 20 & 1.17 \\
     7.98   &  8 & 96$\times$32$\times$32 & 20 & 1.20 \\
     5.03   &  7 & 96$\times$32$\times$32 & 20 & 1.23 \\
     3.18   &  7 & 80$\times$32$\times$32 & 24 & 1.25 \\
     2.00   &  5 & 80$\times$32$\times$32 & 30 & 1.21 \\
     1.26   &  4 & 64$\times$32$\times$32 & 37 & 1.21 \\
     0.798  &  5 & 64$\times$32$\times$32 & 40 & 1.23 \\
     0.503  &  5 & 64$\times$32$\times$32 & 40 & 1.26 \\
     0.318  &  5 & 64$\times$32$\times$32 & 40 & 1.28 \\
     0.200  & 10 & 64$\times$40$\times$40 & 40 & 1.27 \\
     0.126  & 10 & 64$\times$40$\times$40 & 40 & 1.30 \\
     0.0798 & 14 & 80$\times$40$\times$40 & 40 & 1.26 \\
     0.0503 & 15 & 80$\times$40$\times$40 & 40 & 1.28 \\
    %
    \bottomrule
  \end{tabular}
\end{minipage}
\end{table}
%

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/fullspace}
  \caption{(a) Frequency- and (b) time-domain results for the homogeneous space
    model. The red lines are the analytical solutions, the blue circles are the
    actually computed responses\deleted{ with \emg3d}, the black dots are the
    interpolated responses, and the dashed black line the obtained time-domain
    response. The errors are clipped for values smaller than 0.1\,\% and bigger
    than 100\,\%.}
  \label{fig:fullspace}
\end{figure*}
%
Most of the computed values stay below a relative error of 1\,\%, our chosen
adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result \replaced{obtained by
transformation of the frequency-domain response}{from \emg3d}, on
top of the red line which is the analytical result. The relative error is
mostly below 1\,\%, except for early times. However, for practical reasons that
is more than enough. Figure~\ref{fig:fullspace-log} shows the same on a
logarithmic scale, with times up to 10 seconds. It clearly shows that if later
times are required, we would need to adjust our Fourier transform parameters.
%
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/fullspace-log}
  \caption{Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic
    scale. To improve later times we would have to compute lower frequencies;
    to improve earlier times we would have to compute more frequencies per
    decade to get a better resolution.}
  \label{fig:fullspace-log}
\end{figure}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.
